Level-Set method on \texttt{p4est}

### 2014-11-12 paper/tex/introduction.tex
Finished with the algorithm section of paper

%!TEX root = draft.tex
\section{Introduction}
The level-set method, originally proposed by Sethian and Osher \cite{Osher;Sethian:88:Fronts-Propagating-w}, is a popular and powerful framework for tracking arbitrary interfaces that undergo complicated topological changes. As a result, the level-set method has wide range of application such as multiphase flows, moving boundary problems, image segmentation, and computer graphics graphics \cite{Osher;Fedkiw:01:Level-Set-Methods:-A,Sethian:99:Level-set-methods-an}. An important feature which makes this method powerful and easy to use is that the location of interface is defined implicitly on an underlying grid. This convenience, however, comes at a price. First, compared to an explicit method, e.g. front tracking \cite{Juric:96:A-Front-Tracking-Met, Tryggvason;Bunner;Esmaeeli;etal:01:A-Front-Tracking-Met}, level-set method is typically less accurate and mass conservation could be a problem; although progress has been made in resolving this issue \cite{Enright;Fedkiw;Ferziger;etal:02:A-Hybrid-Particle-Le}. Second, the level-set function has to be defined in a higher dimensional space compared to the interface. If only the location of interface is needed, the added dimension greatly increases the overall computational cost. One way to avoid this problem is by computing the level-set only close to the interface, e.g. as in the narrow-band level-set method \cite{Adalsteinsson;Sethian:95:A-Fast-Level-Set-Met} or, more recently, by using a hash table to restrict both computation and storage requirements \cite{Brun;Guittet;Gibou:12:A-local-level-set-me}.

@@

Parallel level-set algorithms can be categorized in two groups: 1) Parallel advection algorithms and 2) Parallel reinitialization algorithms. Eulerian advection schemes can easily be parallelized but unfortunately are limited by the CFL condition which could be very restrictive for adaptive grids. Semi-Lagrangian methods combine the unconditional stability of Lagrangian methods and the ease of use of Eulerian grids and have been successfully used for advecting the level-set function on tree-based grids before \cite{Min;Gibou:07:A-second-order-accur}. However, parallelizing the semi-Lagrangian algorithm in a domain decomposition context is not an easy task. The reason for this is twofold: First, depending on the CFL number, the departure points may end up outside the ghost region and in remote processors that are potentially far away. This requires a very dynamic and nonuniform communication pattern which is complicated to implement. For adaptive grids situation is even more complicated due to the asymmetric nature of communication (c.f. \ref{section::parallel algorithms}). Second, load balancing could be an issue for large CFL numbers and nonuniform velocity fields due to clustering of departure points and can considerably restrict the scalability of the algorithm. Both of these problems, of course, could be avoided by choosing $\text{CFL} \le 1$ but that would defeat the purpose of using semi-Lagrangian algorithm in the first place.

@@

The rest of this article is organized as follows: In section \ref{section::levelset method} we briefly review the sequential algorithms and discretization methods for the level-set equation on adaptive tree-based grids. These ideas are then extended in section \ref{section::parallel algorithms} to parallel using a domain decomposition method. In section \ref{section::scaling results} we provide several examples that illustrate the scalability of our algorithms. Finally, we close by providing an important application of our algorithms in modeling the solidification process by solving a Stefan problem in \ref{section::applications}.

>>>>

%!TEX root = draft.tex
\section{Introduction}\label{sec:introduction}
The level-set method, originally proposed by Sethian and Osher \cite{Osher;Sethian:88:Fronts-Propagating-w}, is a popular and powerful framework for tracking arbitrary interfaces that undergo complicated topological changes. As a result, the level-set method has wide range of application such as multiphase flows, moving boundary problems, image segmentation, and computer graphics graphics \cite{Osher;Fedkiw:01:Level-Set-Methods:-A,Sethian:99:Level-set-methods-an}. An important feature which makes this method powerful and easy to use is that the location of interface is defined implicitly on an underlying grid. This convenience, however, comes at a price. First, compared to an explicit method, e.g. front tracking \cite{Juric:96:A-Front-Tracking-Met, Tryggvason;Bunner;Esmaeeli;etal:01:A-Front-Tracking-Met}, level-set method is typically less accurate and mass conservation could be a problem; although progress has been made in resolving this issue \cite{Enright;Fedkiw;Ferziger;etal:02:A-Hybrid-Particle-Le}. Second, the level-set function has to be defined in a higher dimensional space compared to the interface. If only the location of interface is needed, the added dimension greatly increases the overall computational cost. One way to avoid this problem is by computing the level-set only close to the interface, e.g. as in the narrow-band level-set method \cite{Adalsteinsson;Sethian:95:A-Fast-Level-Set-Met} or, more recently, by using a hash table to restrict both computation and storage requirements \cite{Brun;Guittet;Gibou:12:A-local-level-set-me}.

@@

Parallel level-set algorithms can be categorized in two groups: 1) Parallel advection algorithms and 2) Parallel reinitialization algorithms. Eulerian advection schemes can easily be parallelized but unfortunately are limited by the CFL condition which could be very restrictive for adaptive grids. Semi-Lagrangian methods combine the unconditional stability of Lagrangian methods and the ease of use of Eulerian grids and have been successfully used for advecting the level-set function on tree-based grids before \cite{Min;Gibou:07:A-second-order-accur}. However, parallelizing the semi-Lagrangian algorithm in a domain decomposition context is not an easy task. The reason for this is twofold: First, depending on the CFL number, the departure points may end up outside the ghost region and in remote processors that are potentially far away. This requires a very dynamic and nonuniform communication pattern which is complicated to implement. For adaptive grids situation is even more complicated due to the asymmetric nature of communication (c.f. \ref{sec:parallel algorithms}). Second, load balancing could be an issue for large CFL numbers and nonuniform velocity fields due to clustering of departure points and can considerably restrict the scalability of the algorithm. Both of these problems, of course, could be avoided by choosing $\text{CFL} \le 1$ but that would defeat the purpose of using semi-Lagrangian algorithm in the first place.

@@

The rest of this article is organized as follows: In section \ref{sec:levelset method} we briefly review the sequential algorithms and discretization methods for the level-set equation on adaptive tree-based grids. These ideas are then extended in section \ref{sec:parallel algorithms} to parallel using a domain decomposition method. In section \ref{section::scaling results} we provide several examples that illustrate the scalability of our algorithms. Finally, we close by providing an important application of our algorithms in modeling the solidification process by solving a Stefan problem in \ref{section::applications}.

### 2014-11-12 paper/tex/levelset_method.tex
Finished with the algorithm section of paper

%!TEX root = draft.tex
\section{The level-set method}

\indent The level-set method is a method to capture interfaces implicitely by constructing a function $\phi$, called the level-set function, that is negative on on side of the domain $\Omega^-$ and positive on the other side $\Omega^+$, thus defining the interface $\Gamma$ as the zero contour of $\phi$, $\Gamma = \{\mathbf{x} \vert \phi(\mathbf{x}=0\}$.\\

If infinitely many functions satisfy these requirements, the most natural and most convenient representation is a signed distance function to the interface. The level-set function is therefore reinitialized to a signed distance function at every time step by solving the reinitialization equation

\begin{equation} \label{eq::reinitialization}
\pd{\phi}{\tau} + \mathrm{sign}(\phi) \left( \lvert \phi \rvert - 1 \right) = 0,
\end{equation}

where $\tau$ is a fictitious time. The discretization and implementation of this procedure will be discussed later in section \ref{section::reinitialization}. Geometrical quantities such as the normal to the interface and the local curvature can be computed straightforwardly from the level-set function as

\begin{equation*}
\mathbf{n} = \frac{\nabla \phi}{\lvert \nabla \phi \rvert},
\end{equation*}

where it should be noted that $\lvert \nabla \phi \rvert = 1$ when the level-set function is reinitialized, and

\begin{equation*}
\kappa = \nabla \cdot \mathbf{n} =  \nabla \cdot \left( \frac{\nabla \phi}{\lvert \nabla \phi \rvert} \right).
\end{equation*}

\section{The \texttt{p4est} library}

The work we present relies on the parallel octree library \texttt{p4est} \cite{burstedded:2011:p4est}. It is scalable implementation of the general octree structure in a massively parallel mpi environment. We limit our presentation of \texttt{p4est} to the relevant information for our application, for more details the reader is invited to read the original article \cite{burstedded:2011:p4est}.

The \texttt{p4est} library provides the geometrical information for the leaves of an octree which includes the neighboring information for the leaves (called quadrants in the original article) of the tree, a layer of ghost quadrants and ghost vertices for each process, the coarsening and refining procedures, and encapsulates the communication between processes. However, it does not provide the vertical structure of the octrees. If it is generally unnecessary, it is a desirable feature for a finite differences based code as it speeds up the access to arbitrary quadrants information dramatically. Thus, we choose to reconstruct the local vertical structure of the octree on every processes. Note that given the local and ghost layer information, this is an entirely local procedure and no communication is needed.

In addition to the owned and ghost categories, we choose to group the vertices in \textit{local} and \textit{layer} sets. We define the layer vertices as the vertices that are part of another process's ghost layer and the local vertices as the vertices that are not part of any other process's ghost layer. Figure \ref{fig::local_layer_vertices} gives a graphical representation of a boundary between two processes and the classification of the various quadrants and vertices. The idea behind this sorting is that layer vertices are needed by other processes for finite difference calculations and therefore require to be synchronized while the local vertices are used only by the process they belong to. We make use of this classification to implement scalable algorithms for the level-set method, as explained in the following sections.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=.4\textwidth]{pictures/layer_nodes.pdf}
\caption{Illustration of a possible layout at the interface between two processes. The red quarants and the orange and purples vertices belong to process 1 while the green quadrants and the green and black vertices belong to process 2. The green G denotes the ghost quadrants for process 2, the green dots are the ghost vertices for process 1, and the purple dots correspond to layer vertices for process 1 (and therefore ghost vertices for process 2).} \label{fig::local_layer_vertices}
\end{center}
\end{figure}

>>>>

%!TEX root = draft.tex
\section{The level-set method}\label{sec:levelset method}
The level-set method, introduced in \cite{Osher;Sethian:88:Fronts-Propagating-w}, is an implicit framework for tracking interfaces that undergo complicated topological changes. In this framework an interface is represented by the zero contour of a higher dimensional function, e.g. a curve in two spatial dimension can be described as $\Gamma = \{(x,y) | \phi(x,y) = 0\}$ where $\phi(x,y)$ is the level-set function. Evolution of the curve under a velocity field $\mathbf{V}$ is then obtained by solving the level-set equation:
\be
\phi_t + \mathbf{V} \cdot \nabla \phi = 0.
\label{eq:ls}
\ee
When velocity field does not depend on the level set function itself, equation \eqref{eq:ls} can be solved using the Semi-Lagrangian method. An important advantage of the Semi-Lagrangian method over regular finite difference method is its unconditional stability which allows for arbitrarily large time steps. This is specially of interest when using adaptive grids that allow for higher grid resolutions.

In general, an infinite number of level-set functions can describe the same zero contour and thus interface. However, it is desirable to chose functions with signed distance property, i.e. $|\nabla \phi| = 1$. As detailed in section \ref{sec:introduction}, we solve the pseudo-time transient reinitialization equation \cite{Sussman;Smereka;Osher:94:A-Level-Set-Approach, Osher;Fedkiw:01:Level-Set-Methods:-A} to achieve this property,
\be
\phi_\tau + S(\phi_0)\left(|\nabla \phi| - 1\right) = 0,
\label{eq:reinitialization}
\ee
where $\phi_0$ is any level-set function that correctly describes the interface location and $S(\phi_0)$ is an appropriate approximation of the sign function \cite{Osher;Fedkiw:02:Level-Set-Methods-an}. Here, we do not go into the details of the sequential algorithms for solving equations \eqref{eq:ls} and \eqref{eq:reinitialization}. Instead, we note that the algorithms presented in section \ref{sec:parallel algorithms} are based on the sequential methods presented in \cite{Min;Gibou:07:A-second-order-accur} and refer the interested reader to the aforementioned article and references therein for more details.

% \indent The level-set method is a method to capture interfaces implicitely by constructing a function $\phi$, called the level-set function, that is negative on on side of the domain $\Omega^-$ and positive on the other side $\Omega^+$, thus defining the interface $\Gamma$ as the zero contour of $\phi$, $\Gamma = \{\mathbf{x} \vert \phi(\mathbf{x}=0\}$.\\

% If infinitely many functions satisfy these requirements, the most natural and most convenient representation is a signed distance function to the interface. The level-set function is therefore reinitialized to a signed distance function at every time step by solving the reinitialization equation

% \begin{equation} \label{eq::reinitialization}
% \pd{\phi}{\tau} + \mathrm{sign}(\phi) \left( \lvert \phi \rvert - 1 \right) = 0,
% \end{equation}

% where $\tau$ is a fictitious time. The discretization and implementation of this procedure will be discussed later in section \ref{section::reinitialization}. Geometrical quantities such as the normal to the interface and the local curvature can be computed straightforwardly from the level-set function as

% \begin{equation*}
% \mathbf{n} = \frac{\nabla \phi}{\lvert \nabla \phi \rvert},
% \end{equation*}

% where it should be noted that $\lvert \nabla \phi \rvert = 1$ when the level-set function is reinitialized, and

% \begin{equation*}
% \kappa = \nabla \cdot \mathbf{n} =  \nabla \cdot \left( \frac{\nabla \phi}{\lvert \nabla \phi \rvert} \right).
% \end{equation*}

% \section{The \texttt{p4est} library}

% The work we present relies on the parallel octree library \texttt{p4est} \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}. It is scalable implementation of the general octree structure in a massively parallel mpi environment. We limit our presentation of \texttt{p4est} to the relevant information for our application, for more details the reader is invited to read the original article \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

% The \texttt{p4est} library provides the geometrical information for the leaves of an octree which includes the neighboring information for the leaves (called quadrants in the original article) of the tree, a layer of ghost quadrants and ghost vertices for each process, the coarsening and refining procedures, and encapsulates the communication between processes. However, it does not provide the vertical structure of the octrees. If it is generally unnecessary, it is a desirable feature for a finite differences based code as it speeds up the access to arbitrary quadrants information dramatically. Thus, we choose to reconstruct the local vertical structure of the octree on every processes. Note that given the local and ghost layer information, this is an entirely local procedure and no communication is needed.

% In addition to the owned and ghost categories, we choose to group the vertices in \textit{local} and \textit{layer} sets. We define the layer vertices as the vertices that are part of another process's ghost layer and the local vertices as the vertices that are not part of any other process's ghost layer. Figure \ref{fig::local_layer_vertices} gives a graphical representation of a boundary between two processes and the classification of the various quadrants and vertices. The idea behind this sorting is that layer vertices are needed by other processes for finite difference calculations and therefore require to be synchronized while the local vertices are used only by the process they belong to. We make use of this classification to implement scalable algorithms for the level-set method, as explained in the following sections.

% \begin{figure}[ht!]
% \begin{center}
% \includegraphics[width=.4\textwidth]{pictures/layer_nodes.pdf}
% \caption{Illustration of a possible layout at the interface between two processes. The red quarants and the orange and purples vertices belong to process 1 while the green quadrants and the green and black vertices belong to process 2. The green G denotes the ghost quadrants for process 2, the green dots are the ghost vertices for process 1, and the purple dots correspond to layer vertices for process 1 (and therefore ghost vertices for process 2).} \label{fig::local_layer_vertices}
% \end{center}
% \end{figure}
